---
title: "DATA624 Project 2"
author: "Olivia Azevedo, Ahmed Hassan, Ariba Razzaq, Tatyana Titova"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: "hide"
    theme: flatly
    toc: true
    toc_float:
      collapsed: true
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(fpp3)
library(kableExtra)
library(ggplot2)
library(rio)
library(skimr)
library(zoo)
library(writexl)
library(readxl)
library(caret)
library(dplyr)
library(corrplot)
library(nnet)
library(randomForest)
library(fastDummies)

knit_table <- function(df, caption='', position = 'left', color = '#2d3c52') {
  kbl(df, format = "html", escape = FALSE, caption = caption) |>
    kable_styling(
      full_width = F,
      position = position,
      bootstrap_options = c("striped")) |>
    row_spec(0, bold = T,
             color = "white",
             background = color,
             extra_css = paste("border: 2px solid", color, ";")) |>
    row_spec(dim(df)[1], extra_css = paste("border-bottom: 2px solid", color, ";")) |>
    column_spec(1, extra_css = paste("border-left: 2px solid", color, ";")) |>
    column_spec(dim(df)[2], extra_css = paste("border-right: 2px solid", color, ";")) |>
    scroll_box(width = "100%", height = ifelse(((nrow(df)+1)*45)+50 > 450, "450px", paste0(((nrow(df)+1)*45)+50, "px")), box_css = "border: 1px solid #FFFFFF;")
}
```

# Import Data
```{r}
# import data from github
train_data_url = "https://github.com/okazevedo90/DATA624/raw/refs/heads/main/project_2/StudentData.xlsx"
test_data_url = "https://github.com/okazevedo90/DATA624/raw/refs/heads/main/project_2/StudentEvaluation.xlsx"

train_data = import(train_data_url)
test_data = import(test_data_url)

head(train_data)
head(test_data)
```

# EDA

## Initial Data Exploration
```{r}
# =========================================================
# INITIAL DATA EXPLORATION 
# =========================================================

# View dataset structure
str(train_data)

# Summary statistics
summary(train_data)

# Dataset dimensions
dim(train_data)

# Detailed overview
skim(train_data)
```

## Check Missing Value
```{r}
# =========================================================
# CHECK MISSING VALUES
# =========================================================

# Count missing values in each column
missing_values <- colSums(is.na(train_data))

# Display missing value counts
missing_values

# Create missing value table
missing_table <- data.frame(
  Variable = names(missing_values),
  Missing_Count = missing_values
)

missing_table
```

## Histograms for Numeric Variables
```{r}
# =========================================================
#  HISTOGRAMS FOR NUMERIC VARIABLES
# =========================================================

# Select numeric columns only
numeric_cols <- train_data %>%
  select(where(is.numeric))

# Create histograms for all numeric variables
# for(col in names(numeric_cols)){
#   print(
#     ggplot(train_data, aes(x = .data[[col]])) +
#       geom_histogram(
#         bins = 30,
#         fill = "skyblue",
#         color = "black"
#       ) +
#       labs(
#         title = paste("Histogram of", col),
#         x = col,
#         y = "Frequency"
#       ) +
#       theme_minimal()
#   )
# }


train_data %>%
  select(where(is.numeric)) %>%
  tidyr::pivot_longer(everything(), names_to = "variable", values_to = "value") %>%
  ggplot(aes(x = value)) +
  geom_histogram(bins = 30, fill = "skyblue", color = "black") +
  facet_wrap(~ variable, scales = "free") +
  labs(title = "Histograms of All Numeric Variables") +
  theme_minimal()
```

## Outlier Detection
```{r}

# =========================================================
#  BOXPLOTS FOR OUTLIER DETECTION
# =========================================================

for(col in names(numeric_cols)){
  print(
    ggplot(train_data, aes(y = .data[[col]])) +
      geom_boxplot(fill = "tomato") +
      labs(
        title = paste("Boxplot of", col),
        y = col
      ) +
      theme_minimal()
  )
}

train_data %>%
  select(where(is.numeric)) %>%
  tidyr::pivot_longer(everything(), names_to = "variable", values_to = "value") %>%
  ggplot(aes(y = value)) +
  geom_boxplot(fill = "tomato") +
  facet_wrap(~ variable, scales = "free") +
  labs(title = "Boxplots of All Numeric Variables") +
  theme_minimal()
```

## Correlation Analysis
```{r}
# =========================================================
#  CORRELATION ANALYSIS
# =========================================================

# Compute correlation matrix
cor_matrix <- cor(
  numeric_cols,
  use = "pairwise.complete.obs"
)

# Correlation heatmap
corrplot(
  cor_matrix,
  method = "color",
  type = "upper",
  tl.cex = 0.7
)
```

## Scatterplots 
```{r}
# =========================================================
#  SCATTERPLOTS VS PH
# =========================================================

# Remove PH from predictor list
predictor_cols <- setdiff(names(numeric_cols), "PH")

# Scatterplots against PH
# for(col in predictor_cols){
#   print(
#     ggplot(train_data, aes(x = .data[[col]], y = PH)) +
#       geom_point(alpha = 0.6, color = "blue") +
#       geom_smooth(method = "lm", se = FALSE, color = "red") +
#       labs(
#         title = paste(col, "vs PH"),
#         x = col,
#         y = "PH"
#       ) +
#       theme_minimal()
#   )
# }

train_data %>%
  select(where(is.numeric)) %>%
  tidyr::pivot_longer(-PH, names_to = "variable", values_to = "value") %>%
  ggplot(aes(x = value, y = PH)) +
  geom_point(alpha = 0.3, color = "blue") +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  facet_wrap(~ variable, scales = "free_x") +
  labs(title = "Predictors vs PH") +
  theme_minimal()


```

# Data Prep

## Fearture Engineering
```{r}
# =========================================================
#  FEATURE ENGINEERING
# =========================================================

# Convert Brand Code to categorical variable
if("Brand Code" %in% names(train_data)) {

  train_data$`Brand Code` <- as.character(
    train_data$`Brand Code`
  )

  train_data$`Brand Code`[
    is.na(train_data$`Brand Code`)
  ] <- "Unknown"

  train_data$`Brand Code` <- as.factor(
    train_data$`Brand Code`
  )

  test_data$`Brand Code` <- as.character(
    test_data$`Brand Code`
  )

  test_data$`Brand Code`[
    is.na(test_data$`Brand Code`)
  ] <- "Unknown"

  test_data$`Brand Code` <- factor(
    test_data$`Brand Code`,
    levels = levels(train_data$`Brand Code`)
  )
}
```

```{r}
train_data = dummy_cols(train_data, select_columns = "Brand Code", remove_selected_columns = TRUE)
test_data = dummy_cols(test_data, select_columns = "Brand Code", remove_selected_columns = TRUE)
```



## Seperate Target Variable
```{r}
# =========================================================
# SECTION 10 — SEPARATE TARGET VARIABLE
# =========================================================

# PH is the prediction target
y <- train_data$PH

# Remove PH from predictors
x <- train_data %>%
  select(-PH)
```

## Remove Near-Zero Variance Variables
```{r}
# =========================================================
#  REMOVE NEAR-ZERO VARIANCE VARIABLES
# =========================================================

# Identify low-information predictors
nzv <- nearZeroVar(x)

# Remove them if any exist
if(length(nzv) > 0){

  x <- x[, -nzv]
}

# Match columns in test dataset
test_processed <- test_data[, names(x)]
```

## Remove Highly Corrrelated Variables
```{r}
# =========================================================
#  REMOVE HIGHLY CORRELATED VARIABLES
# =========================================================

# Keep numeric variables only
numeric_x <- x %>%
  select(where(is.numeric))

# Correlation matrix
cor_matrix <- cor(
  numeric_x,
  use = "pairwise.complete.obs"
)

# Identify highly correlated variables
highCorr <- findCorrelation(
  cor_matrix,
  cutoff = 0.90
)

# Remove correlated predictors
if(length(highCorr) > 0){

  highCorr_names <- colnames(numeric_x)[highCorr]

  x <- x[, !(names(x) %in% highCorr_names)]

  test_processed <- test_processed[, names(x)]
}

```



## Data Preprocessing
```{r}
# =========================================================
#  PREPROCESSING
# =========================================================

# Apply:
# 1. Median imputation
# 2. Centering
# 3. Scaling

preprocess_model <- preProcess(
  x,
  method = c(
    "medianImpute",
    "center",
    "scale"
  )
)

# Apply preprocessing to training data
x_processed <- predict(
  preprocess_model,
  x
)

# Apply SAME preprocessing to test data
test_processed <- predict(
  preprocess_model,
  test_processed
)
```

## Final Data Checks
```{r}
# =========================================================
#  FINAL DATASET CHECKS
# =========================================================

# Check for remaining missing values
colSums(is.na(x_processed))

# Final dataset dimensions
dim(x_processed)

# Preview processed dataset
head(x_processed)
```


## Create and Export Final Modeling Dataset
```{r}
# =========================================================
#  CREATE FINAL MODELING DATASET
# =========================================================

# Combine target variable with processed predictors
final_train_data <- cbind(
  PH = y,
  x_processed
)

# Preview final dataset
head(final_train_data)
```


```{r}
# =========================================================
#  EXPORT FINAL DATASETS
# =========================================================

# Export processed training dataset
write_xlsx(
  final_train_data,
  "Processed_Training_Data.xlsx"
)

# Export processed test dataset
write_xlsx(
  test_processed,
  "Processed_Test_Data.xlsx"
)
```


```{r}
final_train_data <- final_train_data[!is.na(final_train_data$PH), ]
```


## Split Data
```{r}
set.seed(123)

# split data into train and test sets
trainingRows <- createDataPartition(final_train_data$PH, p = .80, list= FALSE)

# train 
train <- final_train_data[trainingRows, ] 
# train_x <- final_train_data[trainingRows, ] |> select(-PH)
# train_y <- final_train_data[trainingRows, 'PH']

# test
test = final_train_data[-trainingRows, ]
# test_x <- final_train_data[-trainingRows, ] |> select(-PH)
# test_y <- final_train_data[-trainingRows, 'PH']
```

# Model Development {.tabset}

## Linear Regression
```{r}
set.seed(123)

# 10-fold cross-validation
ctrl <- trainControl(method = "cv", number = 5)

# fit model
lmModel <- train(PH ~ .,
    data = train,
    method = "lm", trControl = ctrl) 

# review performance metrics
lm_pred <- predict(lmModel,  test)
lm_results = postResample(pred = lm_pred, obs = test$PH)
lm_results
```

## Partial Least Squares
```{r}
set.seed(123)

# fit model
plsModel <- train(PH ~ .,data = train,
                 method = "pls", 
                 tuneLength = 20,
                 trControl = ctrl,
                 # preProc = c("center", "scale"),
                 validation = "CV")

# review performance metrics
pls_pred <- predict(plsModel, test)
pls_results = postResample(pls_pred, test$PH)
pls_results
```

## Neural Network
```{r, warning=FALSE}
# remove predictors to ensure maximum abs pairwise corr between predictors < 0.75
tooHigh <-findCorrelation(cor(train_x[, -1]), cutoff = .75)

train_x_nnet <- train_x[, -tooHigh]
test_x_nnet <- test_x[, -tooHigh]

nnetGrid <- expand.grid(
  .decay = c(0, 0.01, .1),.size = c(1:5),
  .bag = FALSE)

# fit model
nnetTune <- train(PH ~ .,data = train,
                  method = "avNNet",
                  tuneGrid = nnetGrid,
                  trControl = ctrl,
                  # preProc = c("center", "scale"),
                  linout = TRUE,
                  trace = FALSE,
                  MaxNWts =  5 * (ncol(train_x_nnet) + 1) + 5 + 1,
                  maxit = 50)

# review performance metrics
nnetPred <- predict(nnetTune, newdata = test)
nnet_results = postResample(nnetPred, test$PH)
nnet_results
```

## MARS
```{r}
marsGrid <- expand.grid(.degree = 1:2, .nprune = 2:38)
# Fix the seed so that the results can be reproduced
set.seed(123)

# fit model
marsTuned <- train(PH ~ .,data = train,
  method = "earth",
  # Explicitly declare the candidate models to test
  tuneGrid = marsGrid,
  trControl = ctrl
  )

# review performance metrics
marsPred <- predict(marsTuned, netest)
mars_results = postResample(marsPred, test$PH)
mars_results
```

## SVM
```{r}
# svm <- train(train_x[-1], train_y,
#   method = "svmRadial",
#   preProc = c("center", "scale"),
#   tuneLength = 14,
#   trControl = trainControl(method = "cv"))
# 
# # review performance metrics
# svmPred <- predict(svm, newdata = test_x)
# svm_results = postResample(svmPred, test_y)
# svm_results
```


## Random Forest
```{r}
set.seed(123)
rfGrid <- expand.grid(mtry = c(2,4,6,8,10))

# fit model
rfModel <- train(PH ~ .,data = train,
                 trControl = ctrl,
                 # tuneGrid = rfGrid,
                 method = "rf")

# review performance metrics
rf_pred <- predict(rfModel, test)
rf_results = postResample(pred = rf_pred, obs = test$PH)
rf_results
```


## Boosted Tree
```{r}
gbmGrid <- expand.grid(interaction.depth = seq(1, 7, by = 2),
                       n.trees = seq(100, 1000, by = 50),
                       shrinkage = c(0.01, 0.1),
                       n.minobsinnode = 10)

set.seed(123)

# fit model
gbm <- train(PH ~ .,data = train,
                 method = "gbm",
                 tuneGrid = gbmGrid,
                 verbose = FALSE)

# review performance metrics
gbm_pred <- predict(gbm, test)
gbm_results = postResample(gbm_pred, obs = test$PH)
gbm_results
```

## Cubist
```{r}
set.seed(123)

# fit model
cubist_model <- train(PH ~ .,data = train,
  method = "cubist",
  trControl = ctrl,
  tuneGrid = expand.grid(committees = c(1, 5, 10), neighbors = c(0, 3, 5, 7))
)

# rownames(cubist_model$trainingData)


# review performance metrics
cubist_pred <- predict(cubist_model, test)
cubist_results = postResample(cubist_pred, test$PH)
cubist_results
```

# Model Evaluation
```{r}
# compile all model performance results
final_results = rbind(
  lm_results,
  mars_results,
  pls_results,
  nnet_results,
  rf_results,
  gbm_results,
  cubist_results
)
# add model name column
final_results = data.frame(cbind(Model = c(
  'Linear Regression',
  'MARS',
  'Partial Least Squares',
  'Neural Network',
  'Random Forest',
  'Boosted Tree',
  'Cubist'
  ), final_results))
rownames(final_results) <- NULL
final_results = final_results |> arrange(RMSE)
# output results to table
knit_table(final_results, 'Model Performance Metrics')

# visualize results
final_results = final_results |>
  mutate(RMSE = as.numeric(RMSE)) |>
  mutate(RMSE = round(as.numeric(RMSE), 2))
ggplot(data = final_results, aes(x = reorder(Model, RMSE), y = RMSE)) +
      geom_bar(stat = "identity", fill='#2d3c52') +
      coord_flip() +
      labs(
        title = "Model Comparison: RMSE",
        x = "Model",
        y = "RMSE"
      ) +
      theme_minimal()
```

# Feature Importance
```{r}
varImp(rfModel)
# plot(varImp(cubist_model), top = 10,
#      main = "Cubist Model Variable Importance - Top 10")
# 


# cubist_vi <- 
# varImp(cubist_model$finalModel, scale = FALSE)
# knit_table(rownames_to_column(cubist_vi, var = "Predictor") |> arrange(desc(Overall)), 'Cubist Variable Importance')
```


# Final Predictions
```{r}
final_pred <- predict(cubist, test_processed)
final_pred
```


