---
title: "DATA624 Project 2"
author: "Olivia Azevedo, Ahmed Hassan, Ariba Razzaq, Tatyana Titova"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: "hide"
    theme: flatly
    toc: true
    toc_float:
      collapsed: true
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(fpp3)
library(kableExtra)
library(ggplot2)
library(rio)
library(skimr)
library(zoo)
library(writexl)
library(readxl)
library(caret)
library(dplyr)
library(corrplot)
library(nnet)
library(randomForest)
library(fastDummies)

knit_table <- function(df, caption='', position = 'left', color = '#2d3c52') {
  kbl(df, format = "html", escape = FALSE, caption = caption) |>
    kable_styling(
      full_width = F,
      position = position,
      bootstrap_options = c("striped")) |>
    row_spec(0, bold = T,
             color = "white",
             background = color,
             extra_css = paste("border: 2px solid", color, ";")) |>
    row_spec(dim(df)[1], extra_css = paste("border-bottom: 2px solid", color, ";")) |>
    column_spec(1, extra_css = paste("border-left: 2px solid", color, ";")) |>
    column_spec(dim(df)[2], extra_css = paste("border-right: 2px solid", color, ";")) |>
    scroll_box(width = "100%", height = ifelse(((nrow(df)+1)*45)+50 > 450, "450px", paste0(((nrow(df)+1)*45)+50, "px")), box_css = "border: 1px solid #FFFFFF;")
}

model_perf_df = function(model, test_df, model_name) {
  pred <- predict(model, test_df)
  results = postResample(pred = pred, obs = test_df$PH)
  results_df = rownames_to_column(data.frame(results), var = "Metric")
  colnames(results_df) = c('Metric', 'Score')
  knit_table(results_df, paste0(model_name, ' Performance'))
  return(results)
}

load_models = TRUE
```

## Import Data
```{r import}
# import data from github
train_data_url = "https://github.com/okazevedo90/DATA624/raw/refs/heads/main/project_2/Data/StudentData.xlsx"
test_data_url = "https://github.com/okazevedo90/DATA624/raw/refs/heads/main/project_2/Data/StudentEvaluation.xlsx"

train_data = import(train_data_url)
test_data = import(test_data_url)

knit_table(head(train_data), 'View Head of train data')
knit_table(head(test_data), 'View Head of test data')
```

<hr style="border:2px solid #2d3c52">

## EDA

### Initial Data Exploration
```{r data-exploration}
# View dataset structure
str(train_data)

# Summary statistics
# summary(train_data)

# Dataset dimensions
knit_table(
  tibble(
    Dimmension = c('Rows', 'Columns'),
    Count = dim(train_data))
)

# Detailed overview
skim(train_data)
```

### Check Missing Values
```{r missing-values}
# Count missing values in each column
missing_table <- rownames_to_column(data.frame(colSums(is.na(train_data))), 'Variable')
colnames(missing_table) = c('Variable',  'Missing_Count')

knit_table(missing_table, 'Column Missing Value Count')
```

### Histograms for Numeric Variables
```{r histograms, fig.show="hold", out.width="50%", warning=FALSE}
# Select numeric columns only
numeric_cols <- train_data %>%
  select(where(is.numeric))

# Create histograms for all numeric variables
for(col in names(numeric_cols)){
  print(
    ggplot(train_data, aes(x = .data[[col]])) +
      geom_histogram(
        bins = 30,
        fill = "#00BFC4",
        color = "black"
      ) +
      labs(
        title = paste("Histogram of", col),
        x = col,
        y = "Frequency"
      ) +
      theme_minimal() +
      theme(text = element_text(size = 16))
  )
}
```

### Outlier Detection
```{r outlier, fig.show="hold", out.width="50%", warning=FALSE, fig.height=2}
for(col in names(numeric_cols)){
  print(
    ggplot(train_data, aes(y = .data[[col]])) +
      geom_boxplot(fill = "#22a298", outlier.colour = "coral1") +
      labs(
        title = paste("Boxplot of", col),
        y = col
      ) +
      coord_flip() +
      theme_minimal() +
      theme(text = element_text(size = 16))
  )
}
```

### Correlation Analysis
```{r correlation-analysis, fig.height=12, fig.width=12}
# Compute correlation matrix
cor_matrix <- cor(
  numeric_cols,
  use = "pairwise.complete.obs"
)

# Correlation heatmap
corrplot(
  cor_matrix,
  type = 'lower', order = 'hclust', tl.col = 'black',
         cl.ratio = 0.2, tl.srt = 45, col = COL2('RdYlBu', 10))
```

### Scatterplots
```{r scatterplots, fig.show="hold", out.width="50%", warning=FALSE}
# Remove PH from predictor list
predictor_cols <- setdiff(names(numeric_cols), "PH")

# Scatterplots against PH
for(col in predictor_cols){
  print(
    ggplot(train_data, aes(x = .data[[col]], y = PH)) +
      geom_point(alpha = 0.6, color = "#00BFC4") +
      geom_smooth(method = "lm", formula = 'y ~ x', se = FALSE, color = "coral1") +
      labs(
        title = paste(col, "vs PH"),
        x = col,
        y = "PH"
      ) +
      theme_minimal() +
      theme(text = element_text(size = 16))
  )
}
```

<hr style="border:2px solid #2d3c52">

## Data Prep

### Fearture Engineering
```{r feature-engineering}
# Convert Brand Code to categorical variable
if("Brand Code" %in% names(train_data)) {

  train_data$`Brand Code` <- as.character(
    train_data$`Brand Code`
  )

  train_data$`Brand Code`[
    is.na(train_data$`Brand Code`)
  ] <- "Unknown"

  train_data$`Brand Code` <- as.factor(
    train_data$`Brand Code`
  )

  test_data$`Brand Code` <- as.character(
    test_data$`Brand Code`
  )

  test_data$`Brand Code`[
    is.na(test_data$`Brand Code`)
  ] <- "Unknown"

  test_data$`Brand Code` <- factor(
    test_data$`Brand Code`,
    levels = levels(train_data$`Brand Code`)
  )
}

# Transform Brand Code into Dummy Variables
train_data = dummy_cols(train_data, select_columns = "Brand Code", remove_selected_columns = TRUE)
test_data = dummy_cols(test_data, select_columns = "Brand Code", remove_selected_columns = TRUE)
```


### Seperate Target Variable
```{r seperate}
# PH is the prediction target
y <- train_data$PH

# Remove PH from predictors
x <- train_data %>%
  select(-PH)
```

### Remove Variables
#### Remove Near-Zero Variance Variables
```{r low variance}
# Identify low-information predictors
nzv <- nearZeroVar(x)

# Remove them if any exist
if(length(nzv) > 0){

  x <- x[, -nzv]
}

# Match columns in test dataset
test_processed <- test_data[, names(x)]
```

#### Remove Highly Corrrelated Variables
```{r high-corr}
# Keep numeric variables only
numeric_x <- x %>%
  select(where(is.numeric))

# Correlation matrix
cor_matrix <- cor(
  numeric_x,
  use = "pairwise.complete.obs"
)

# Identify highly correlated variables
highCorr <- findCorrelation(
  cor_matrix,
  cutoff = 0.90
)

# Remove correlated predictors
if(length(highCorr) > 0){

  highCorr_names <- colnames(numeric_x)[highCorr]

  x <- x[, !(names(x) %in% highCorr_names)]

  test_processed <- test_processed[, names(x)]
}
```

### Data Preprocessing
```{r preprocessing}
# Apply:
# 1. Median imputation
# 2. Centering
# 3. Scaling

preprocess_model <- preProcess(
  x,
  method = c(
    "medianImpute",
    "center",
    "scale"
  )
)

# Apply preprocessing to training data
x_processed <- predict(
  preprocess_model,
  x
)

# Apply SAME preprocessing to test data
test_processed <- predict(
  preprocess_model,
  test_processed
)
```

### Final Data Checks
```{r final-check}
# Check for remaining missing values
final_missing_table <- rownames_to_column(data.frame(colSums(is.na(x_processed))), 'Variable')
colnames(final_missing_table) = c('Variable',  'Missing_Count')

knit_table(final_missing_table, 'Column Missing Value Count')

# Final dataset dimensions
knit_table(
  tibble(
    Dimmension = c('Rows', 'Columns'),
    Count = dim(x_processed))
)

# Preview processed dataset
knit_table(head(x_processed), 'Preview Processed Dataset')
```

### Final Modeling Dataset
```{r final-data}
# Combine target variable with processed predictors
final_train_data <- cbind(
  PH = y,
  x_processed
)

# Preview final dataset
knit_table(head(final_train_data), 'Preview Final Dataset')

```

```{r export-data}
# Export processed training dataset
write_xlsx(
  final_train_data,
  "Results/Processed_Training_Data.xlsx"
)

# Export processed test dataset
write_xlsx(
  test_processed,
  "Results/Processed_Test_Data.xlsx"
)
```


**UPDATE - AHMED**
```{r}
final_train_data <- final_train_data[!is.na(final_train_data$PH), ]
```


### Split Data
```{r split}
set.seed(123)

# split data into train and test sets
trainingRows <- createDataPartition(final_train_data$PH, p = .80, list= FALSE)

# train
train <- final_train_data[trainingRows, ]

# test
test = final_train_data[-trainingRows, ]
```

<hr style="border:2px solid #2d3c52">

## Model Development {.tabset}

### Linear Regression
```{r linear-regression}
set.seed(123)

# 10-fold cross-validation
ctrl <- trainControl(method = "cv", number = 5)

if (load_models == TRUE) {
  # load model
  lmModel <- readRDS("Models/lmModel.rds")
} else {
  # fit model
  lmModel <- train(PH ~ .,
      data = train,
      method = "lm", trControl = ctrl)
  # save model
  saveRDS(lmModel, "Models/lmModel.rds")
}

# review performance metrics
lm_results = model_perf_df(lmModel, test, 'Linear Regression')
lm_results
```

### Partial Least Squares
```{r pls}
set.seed(123)

if (load_models == TRUE) {
  # load model
  plsModel <- readRDS("Models/plsModel.rds")
} else {
  # fit model
  plsModel <- train(PH ~ .,data = train,
                   method = "pls",
                   tuneLength = 20,
                   trControl = ctrl,
                   validation = "CV")
    # save model
  saveRDS(plsModel, "Models/plsModel.rds")
}

# review performance metrics
pls_results = model_perf_df(plsModel, test, 'Partial Least Squares')
pls_results
```

### Neural Network
```{r, warning=FALSE}
# remove predictors to ensure maximum abs pairwise corr between predictors < 0.75
tooHigh <-findCorrelation(cor(train[, -1]), cutoff = .75)

train_x_nnet <- train[, -tooHigh]
test_x_nnet <- test[, -tooHigh]

nnetGrid <- expand.grid(
  .decay = c(0, 0.01, .1),.size = c(1:5),
  .bag = FALSE)

if (load_models == TRUE) {
  # load model
  nnetTune <- readRDS("Models/nnetTune.rds")
} else {
  # fit model
  nnetTune <- train(PH ~ .,data = train_x_nnet,
                    method = "avNNet",
                    tuneGrid = nnetGrid,
                    trControl = ctrl,
                    linout = TRUE,
                    trace = FALSE,
                    MaxNWts =  5 * (ncol(train_x_nnet) + 1) + 5 + 1,
                    maxit = 50)
  # save model
  saveRDS(nnetTune, "Models/nnetTune.rds")
}

# review performance metrics
nnet_results = model_perf_df(nnetTune, test, 'Neural Network')
nnet_results
```

### MARS
```{r mars}
marsGrid <- expand.grid(.degree = 1:2, .nprune = 2:38)
# Fix the seed so that the results can be reproduced
set.seed(123)

if (load_models == TRUE) {
  # load model
  marsTuned <- readRDS("Models/marsTuned.rds")
} else {
  # fit model
  marsTuned <- train(PH ~ .,data = train,
    method = "earth",
    # Explicitly declare the candidate models to test
    tuneGrid = marsGrid,
    trControl = ctrl
    )
  # save model
  saveRDS(marsTuned, "Models/marsTuned.rds")
}

# review performance metrics
mars_results = model_perf_df(marsTuned, test, 'MARS')
mars_results
```

### Random Forest
```{r random-forest}
set.seed(123)
rfGrid <- expand.grid(mtry = c(8, 10, 15, 20))

if (load_models == TRUE) {
  # load model
  rfModel <- readRDS("Models/rfModel.rds")
} else {
  # fit model
  rfModel <- train(PH ~ .,data = train,
                   tuneGrid = rfGrid,
                   trControl = ctrl,
                   method = "rf")
  # save model
  saveRDS(rfModel, "Models/rfModel.rds")
}

# review performance metrics
rf_results = model_perf_df(rfModel, test, 'Random Forest')
rf_results
```

### Boosted Tree
```{r boosted-tree}
# gbmGrid <- expand.grid(interaction.depth = seq(1, 7, by = 2),
#                        n.trees = seq(100, 1000, by = 50),
#                        shrinkage = c(0.01, 0.1),
#                        n.minobsinnode = 10)
# set.seed(123)
#
# if (load_models == TRUE) {
#   # load model
#   gbm <- readRDS("Models/gbm.rds")
# } else {
#   # fit model
#   gbm <- train(PH ~ .,data = train,
#                  method = "gbm",
#                  tuneGrid = gbmGrid,
#                  verbose = FALSE)
#   # save model
#   saveRDS(gbm, "Models/gbm.rds")
# }
#
# # review performance metrics
# gbm_results = model_perf_df(gbm, test, 'Boosted Tree')
```

### Cubist
```{r cubist}
set.seed(123)


if (load_models == TRUE) {
  # load model
  cubist_model <- readRDS("Models/cubist_model.rds")
} else {
  # fit model
  cubist_model <- train(PH ~ .,data = train,
    method = "cubist",
    trControl = ctrl,
    tuneGrid = expand.grid(committees = c(1, 5, 10), neighbors = c(0, 3, 5, 7))
  )
  # save model
  saveRDS(cubist_model, "Models/cubist_model.rds")
}

# review performance metrics
cubist_results = model_perf_df(cubist_model, test, 'Cubist')
cubist_results
```

<hr style="border:2px solid #2d3c52">

## Model Evaluation

### Model Comparison
```{r model-comparison}
# compile all model performance results
final_results = rbind(
  lm_results,
  mars_results,
  pls_results,
  nnet_results,
  rf_results,
  # gbm_results,
  cubist_results
)

# add model name column
final_results = data.frame(cbind(Model = c(
  'Linear Regression',
  'MARS',
  'Partial Least Squares',
  'Neural Network',
  'Random Forest',
  # 'Boosted Tree',
  'Cubist'
  ), final_results))
rownames(final_results) <- NULL
final_results = final_results |> arrange(RMSE)
# output results to table
knit_table(final_results, 'Model Performance Metrics')

# visualize results
final_results = final_results |>
  mutate(RMSE = as.numeric(RMSE)) |>
  mutate(RMSE = round(as.numeric(RMSE), 3))

ggplot(data = final_results, aes(x = reorder(Model, RMSE), y = RMSE)) +
  geom_bar(stat = "identity", fill='darkolivegreen3') +
  coord_flip() +
  labs(
    title = "Model Comparison: RMSE",
    x = "Model",
    y = "RMSE"
  ) +
  theme_minimal() +
  theme(text = element_text(size = 15))
```

<hr style="border:2px solid #2d3c52">

## Final Model - Random Forest

### Predictor Count Performance
```{r pred-count, warning=FALSE}
ggplot(rfModel) +
  theme_minimal() +
  geom_point(size=2, color="#22a298") +
  labs(
    title = "Random Forest # Predictors by RMSE",
    y = 'RMSE',
    x='# Randomly Selected Predictors'
  )
```

### Feature Importance
```{r feature-imp}
# estimate the importance of each predictor
rf_importance = varImp(rfModel)[1]$importance
rf_importance = rownames_to_column(rf_importance, var = "Predictor")
rf_importance = rf_importance |>
  mutate(Predictor = gsub('`', '', Predictor)) |>
  arrange(desc(Overall))

# visualize results
rf_importance |>
  ggplot(aes(x=reorder(Predictor, Overall), y=Overall)) +
  geom_segment( aes(xend=Predictor, yend=0)) +
  geom_point(size=2, color="#22a298") +
  coord_flip() +
  theme_minimal() +
  labs(
    title = "Random Forest Feature Importance - Top 10",
    y = 'Importance',
    x=''
  )

# output results
knit_table(rf_importance, 'Random Forest Predictor Estimated Importance')
```

<hr style="border:2px solid #2d3c52">

## Final Predictions
```{r final-preds}
# apply model to final test set
final_pred <- predict(rfModel, test_processed)

# export results to xlsx
final_xslx = final_pred |>
  as.data.frame()
colnames(final_xslx) = c('PH_Predictions')

write_xlsx(final_xslx, path = "Results/DATA624_Project2_Predictions.xlsx")

knit_table(final_xslx, 'Final Predictions')
```
